---
title: "find_exemplar"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{find_exemplar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ironr)
```

# Background

One of the first steps in using IRON is the identification of a reference
sample to normalize all other samples to. We use the term `exemplar` as
it is the "example" representing the data. This reference sample could be
determined on first principles. For instance, a stage II breast tumor with no
genomic alterations to normalize against for all breast tumors. More often,
however, we attempt to empirically identify the most central sample from among those
being normalized.

Think of a 50k/100k dimensional space where each sample is represented by a
single point. We would like the `exemplar` of this point cloud; the point that
is most "central" or in the middle of the cloud. That is the purpose of the
original `findmedian` function in libaffy. Normalization generally starts by
identifying this exemplar, then running normalization of all samples against 
the exemplar.

To fully describe the functionality of `findmedian` and equivalent ironr functions,
we first describe the specific approach of *medoids* then generalize this to
`exemplars`, of which `findmedian` is a special case.

## Medoid
The medoid is exactly the concept we looking for when thinking about the representative
sample. Specifically, it appears to be defined as the representative of the data with
the minimum sum of dissimilarities to all other data elements. Note that a medoid
is an element of the data set, whereas centroids are averages of elements and therefore
may not exist as a distinct point in the existing data. 

Some key points:

- The distance metric could be Euclidean or others.
- Finding the medoid requires an all-vs-all distance matrix.

## Exemplars as a Medoid Extension
In this package, we refer to finding an `exemplar` as a more generic approach to
finding the medoid. Specifically, we consider any distance measure for minimizing
and we consider minimizing summarizations other than sums. So in our case, an
exemplar is a representative of the data with the minimum summarization of dissimilarity
to all other data elements. 

### Distance
While we used the term `distance metric` above,
the more general case would be a `distance measure`. There are several characteristics
of a distance metric (see https://en.wikipedia.org/wiki/Metric_space) that are important
for determining the validity (or convergence) of various techniques using it. For
our purposes, we relax these restraints and accept the problematic nature that ensues.
Assymetric distances, etc are all considered acceptable for the more general case.

### Summarization
While the original medoid definition referred to minimizing the sum of distances,
we can also relax this constraint. For instance, rather than the sum we could use
the minimum "mean" distance, minimum "median" distance of points in the data. As
is the case with distance metrics vs. measures, this has deeper implications in the
correctness or assurance that can come from math but allows us a broader overall definition.


## findmedian definition
The original `libaffy` definition of finding the exemplar was known as `findmedian()`.
There are many options that exist when using it, but it appears that the default definition
is:
- rmsd as the distance measure (square root of the mean of squared differences)
- minimize the mean distance

Therefore, rather than choose the exemplar with the minimum sum of Euclidean distances
to all other points in the data, findmedian is defined as choosing the exemplar with
the minimum mean of rmsd distances to all other points in the data.


# Usage
With the background in mind, the `ironr` R package provides support for findmedian,
medoids and finding exemplars. It does this through a set of distance functions (similar
to `stats::dist()`), an R object called `exemplar`, and routines to find the exemplar. 
Each is discussed below.

## Distance
The `stats::dist()` function is incredibly well-optimized for performance. However, it
is limited in specific distance metrics (and measures). As such, we reimplemented the
approach in a more general way since we knew that rmsd (among other distances) would be
required. There is a separate vignette on this part of the packge, but the summary is 
that we support distances by row/column and arbitrary functions for calculating the
distance (in addition to built-in functions). This makes the function super-slow is
large-scale settings but provides the maximal flexibility. Of particular note is that
we implemented an approach for making the Pearson correlation a distance metric (see `pearsonm_distance()`)
as well as a distance correction for missing data (`na.dist_norm()`). These are
integrated into the `distance()` function implemented in this package.

## Exemplar
We created an object called an `exemplar` that is used to hold not only the exemplar,
but also details about how it was found. By way of example, consider the numeric data
in the iris dataset. If we treat each variable (Sepal.Length, Sepal.Width, Petal.Length,
Petal.Width) as "examples", then we could find the medoid variable from these:

```{r}
e <- find_exemplar(iris[,1:4])
```

Note that the `find_exemplar()` function defaults to using the mediod definition and
returns an `exemplar` object. There is a default print statement that shows the
exemplar name, as well as statistics around it and the other options that were not
selected:
```{r}
print(e)
```

In our case, we used the medoid definition so the column with the minimum 
(summed) distance is chosen (e.g. "Sepal.Width"). 

**NOTE**: If you are interested in getting the `medoid` of a dataset, there is
a function `medoid()` which specifically calculates this a bit more efficiently.


The code for these routines is found in `R/exemplar.R`
